---
title: "Powerball Analyses"
author: "Ethan P Heinzen"
date: "8/19/2017"
output: html_document
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 8, message = FALSE, warning = FALSE)
```

```{r}
source("powerball.R")
numbers <- get_data() %>% 
  filter()
```

Below are the published (and easily verifiable) probabilities of winning the Powerball,
with their respective earnings:

```{r results = 'asis'}
odds <- read_csv("Odds.csv", col_names = TRUE, col_types = cols()) %>% 
  mutate(Earnings = replace(Earnings, 1, NA))
e <- sum(odds$Earnings[-1]/odds$one_over_p[-1])
knitr::kable(odds)
```

The expected value of non-jackpot prizes is `r round(e, 4)` dollars.


To calculate the **overall** expected value, 
we'll need to take into account multiple jackpot winners.

```{r plotly = TRUE}
ggplot(numbers, aes(x = Pot, y = Num_Tix)) +
  geom_point(shape = 1) +
  ggtitle("Number of Tickets Sold by Jackpot Size") +
  xlab("Jackpot Size") +
  ylab("Number of Tickets Sold")
```

We take the log-log transform.

```{r}
ggplot(numbers, aes(x = Pot, y = Num_Tix)) +
  scale_y_log10() +
  scale_x_log10(breaks = c(100, 200, 300, 400, 500, 750, 1000, 1500)) +
  geom_point(shape = 1) +
  ggtitle("Number of Tickets Sold by Jackpot Size") +
  xlab("Jackpot Size") +
  ylab("Number of Tickets Sold")
```

We'll try to fit a quadratic model on the log-log scale. We subset to pots at least $40 million.

```{r}
tix.lm <- lm(log(Num_Tix) ~ log(Pot) + I(log(Pot)^2), data = numbers, subset = Pot >= 40)
```

```{r, eval = FALSE}
par(mfrow = c(2,2))
plot(tix.lm)
summary(tix.lm)
par(mfrow = c(1,1))
shapiro.test(residuals(tix.lm))

```

This isn't a particularly great model, especially for the outliers, but we'll go with it.
Our best-fit equation is then:

$ln{(\widehat{NumTix})} = `r coef(tix.lm)[1]` + `r coef(tix.lm)[2]`*ln{(Pot)} + `r coef(tix.lm)[3]`*(ln{(Pot)})^{2}$.

```{r}
pot.df <- data.frame(Pot = seq(40, 1200))
pot.df$Num_Tix <- exp(predict(tix.lm, newdata = pot.df))
ggplot(numbers, aes(x = Pot, y = Num_Tix)) +
  scale_y_log10() +
  scale_x_log10(breaks = c(100, 200, 300, 400, 500, 750, 1000)) +
  geom_point(shape = 1) +
  ggtitle("log(Number of Tickets Sold) by Jackpot Size") +
  xlab("Jackpot Size") +
  ylab("log(Number of Tickets Sold)") +
  geom_line(aes(x = Pot, y = Num_Tix), data = pot.df, color = "red")

ggplot(numbers, aes(x = Pot, y = Num_Tix)) +
  geom_point(shape = 1) +
  ggtitle("Number of Tickets Sold by Jackpot Size") +
  xlab("Jackpot Size") +
  ylab("Number of Tickets Sold") +
  geom_line(aes(x = Pot, y = Num_Tix), data = pot.df, color = "red")
```

We will use the Poisson approximation to the binomial distribution to find the probability
that a given number of people wins the jackpot at a given value (assuming the distribution
of tickets is uniform). In this approximation, when $n \gg 1$ and $p \ll 1$, $\lambda \approx np$.
Here, we have a function $NumTix(Pot) \gg 1$ and $p = \frac{1}{292201338} \ll 1$.
Thus, $\lambda_{Pot} = \frac{NumTix(Pot)}{292201338}$.

```{r, eval = FALSE}
lambda <- (pot.df$Num_Tix)/odds$one_over_p[1]
tmp <- do.call(cbind, lapply(0:6, function(x) dpois(x, lambda))) %>% 
  set_colnames(c("Zero", "One", "Two", "Three", "Four", "Five", "Six"))
seven_or_more <- 1 - rowSums(tmp)
pot.df <- pot.df %>% 
  bind_cols(as.data.frame(tmp)) %>% 
  mutate(Seven_or_More = seven_or_more) %>% 
  mutate(At_Least_One = 1 - Zero)

pot.df.long <- reshape2::melt(pot.df, id.vars = c("Pot", "Num_Tix"))


ggplot(pot.df.long) +
  geom_line(aes(x = Num_Tix, y = value, color = variable)) +
  ggtitle("Probability of Number of Winners at a Given Number of Tickets Sold") +
  ylim(0,1) +
  xlab("Number of Tickets Sold") + ylab("Probability")


ggplot(pot.df.long) +
  scale_x_continuous(breaks = seq(0, 1300, by = 100)) +
  geom_line(aes(x = Pot, y = value, color = variable)) +
  ggtitle("Probability of Number of Winners at a Given Pot") +
  ylim(0,1) +
  xlab("Pot") + ylab("Probability")
```

Side note: because we're using the Poisson distribution, we know that the expected
number of winning tickets for a given pot is also given by
$\lambda_{Pot} = \frac{NumTix(Pot)}{292201338}$.

Let $P_{Pot}(i)$ denote the Poisson probability of $i$ **other** winners when the pot is of size $Pot$.
To calculate the expected value of the pot for **you**, then, we calculate:

$$E(Pot) = \frac{Pot}{1}*P_{Pot}(0) + \frac{Pot}{2}*P_{Pot}(1) + \frac{Pot}{3}*P_{Pot}(2) + ...$$

or

$$E(Pot) = Pot*\sum_{i=0}^{\infty} \frac{P_{Pot}(i)}{i+1}$$

or

$$E(Pot) = Pot*\sum_{i=0}^{\infty} \frac{\lambda_{Pot}^{i} e^{-\lambda_{Pot}}}{(i+1)i!} $$

or (changing indexes)

$$E(Pot) = \frac{Pot}{\lambda_{Pot}}*\sum_{i=1}^{\infty} \frac{\lambda_{Pot}^{i} e^{-\lambda_{Pot}}}{i!}$$

or

$$E(Pot) = \frac{Pot}{\lambda_{Pot}}(1-P_{Pot}(0))$$

Note that this is
the expected value of the pot, not the pot's contribution to your expected
value from buying a ticket. Thus, we get something like this:

$$E_{Pot}(Ticket) = `r e` + \frac{E(Pot)}{292201338}$$.

```{r, eval = FALSE}
pot.df$Expected.Jackpot <- pot.df %>% 
  select(Pot, Num_Tix, Zero) %>% 
  as.matrix() %>% 
  apply(1, function(x) x[1]*1000000*(1 - x[3])/x[2])
pot.df$Expected <- pot.df$Expected.Jackpot + e



pot.df %>% 
  select(Pot, Expected.Jackpot, Expected) %>% 
  reshape2::melt(id.vars = "Pot") %>% 
  ggplot() +
    scale_x_continuous(breaks = seq(0, 1300, by = 100)) +
    scale_y_continuous(breaks = 0.5*(0:8), minor_breaks = seq(0, 4, by = 0.1)) +
    geom_line(aes(x = Pot, y = value, color = variable)) +
    ggtitle("Expectation at a Given Pot") +
    xlab("Pot") +
    ylab("Expectation") +
    theme(legend.position = 'top')
```
