---
title: "Powerball Analyses"
author: "Ethan P Heinzen"
date: "8/19/2017"
output: html_document
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 8, message = FALSE, warning = FALSE)
```

```{r}
source("powerball.R")
numbers <- get_data() %>% 
  filter(!is.na(Num_Tix))
```

Below are the published (and easily verifiable) probabilities of winning the Powerball,
with their respective earnings:

```{r results = 'asis'}
odds <- read_csv("Odds.csv", col_names = TRUE, col_types = cols()) %>% 
  mutate(Earnings = replace(Earnings, 1, NA))
e <- sum(odds$Earnings[-1]/odds$one_over_p[-1])
knitr::kable(odds)
```

The expected value of non-jackpot prizes is `r round(e, 4)` dollars.


To calculate the **overall** expected value, 
we'll need to take into account multiple jackpot winners.

We'll want to only consider recent years, as ticket sales appear to be associated with the
price of the ticket:

```{r plotly = TRUE}
numbers %>% 
  mutate(Price_Per_Tix = paste0("$", Price_Per_Tix)) %>% 
  ggplot(aes(x = Pot, y = Num_Tix, color = Price_Per_Tix)) +
  geom_point(shape = 1) +
  ggtitle("Number of Tickets Sold by Jackpot Size") +
  xlab("Jackpot Size") +
  ylab("Number of Tickets Sold")
```

If we take the log-log transform, it becomes more obvious.

```{r}
numbers %>% 
  mutate(Price_Per_Tix = paste0("$", Price_Per_Tix)) %>% 
  ggplot(aes(x = Pot, y = Num_Tix, color = Price_Per_Tix)) +
  scale_y_log10() +
  scale_x_log10(breaks = c(100, 200, 300, 400, 500, 750, 1000, 1500), minor_breaks = NULL) +
  geom_point(shape = 1) +
  ggtitle("Number of Tickets Sold by Jackpot Size") +
  xlab("Jackpot Size") +
  ylab("Number of Tickets Sold")
```

Among $2 ticket pots, does year associate with ticket sales?

```{r}
numbers <- numbers %>% 
  mutate(
    last_log_Pot = lag(log(Pot)),
    year = lubridate::year(Date),
    year.fct = factor(year)
  ) %>% 
  filter(Price_Per_Tix == 2) %>% 
  mutate(
    new_Pot = Pot == 40,
    jackpot_run = 1 + cumsum(new_Pot),
    last_log_Pot = if_else(new_Pot, 0, last_log_Pot)
  )
ggplot(numbers, aes(x = Pot, y = Num_Tix, color = year)) +
  scale_y_log10() +
  scale_x_log10(breaks = c(100*(1:5), 750, 1000, 1500)) +
  geom_point(shape = 1) +
  ggtitle("Number of $2 Tickets Sold By Jackpot Size") +
  xlab("Jackpot Size") +
  ylab("Number of Tickets Sold")
```

We'll try to fit a quadratic model on the log-log scale. We subset to pots sold with $2 tickets.

```{r}
# tix.lm <- lm(log(Num_Tix) ~ log(Pot) + I(log(Pot)^2), data = numbers)
quants <- log(c(60, 100, 180, 280, 400))
tix.lm <- lm(log(Num_Tix) ~ splines::bs(log(Pot), knots = quants, degree = 1), data = numbers)
tix.lm2 <- lm(log(Num_Tix) ~ splines::bs(log(Pot), knots = quants, degree = 1) + year, data = numbers)
tix.lm3 <- loess(log(Num_Tix) ~ log(Pot), data = numbers)
summary(tix.lm)
```

```{r, eval = FALSE}
par(mfrow = c(2,2))
plot(tix.lm2)
par(mfrow = c(1,1))
shapiro.test(residuals(tix.lm))
```

This isn't a particularly great model for the outliers and low pots, but we'll go with it.

```{r}
pot.df <- expand.grid(Pot = seq(from = 40, to = 1500, by = 1), year = 2012:2019)
pot.df$Num_Tix <- exp(predict(tix.lm2, newdata = pot.df))
ggplot(numbers, aes(x = Pot, y = Num_Tix, color = factor(year))) +
  scale_y_log10() +
  scale_x_log10(breaks = c(100, 200, 300, 400, 500, 750, 1000)) +
  geom_point(shape = 1) +
  ggtitle("Number of Tickets Sold by Jackpot Size") +
  xlab("Jackpot Size") +
  ylab("Number of Tickets Sold") +
  geom_line(data = pot.df, aes(group = year))

ggplot(numbers, aes(x = Pot, y = Num_Tix, color = factor(year))) +
  geom_point(shape = 1) +
  ggtitle("Number of Tickets Sold by Jackpot Size") +
  xlab("Jackpot Size") +
  ylab("Number of Tickets Sold") +
  geom_line(data = pot.df, aes(group = year))
```

We will use the Poisson approximation to the binomial distribution to find the probability
that a given number of people wins the jackpot at a given value (assuming the distribution
of tickets is uniform). In this approximation, when $n \gg 1$ and $p \ll 1$, $\lambda \approx np$.
Here, we have a function $NumTix(Pot) \gg 1$ and $p = \frac{1}{292201338} \ll 1$.
Thus, $\lambda_{Pot} = \frac{NumTix(Pot)}{292201338}$.

Note that we'll use the predictions for the year 2019.

```{r}
pot.df <- filter(pot.df, year == 2019)
lambda <- (pot.df$Num_Tix)/odds$one_over_p[1]
tmp <- do.call(cbind, lapply(0:6, function(x) dpois(x, lambda))) %>% 
  set_colnames(c("Zero", "One", "Two", "Three", "Four", "Five", "Six"))
seven_or_more <- 1 - rowSums(tmp)
pot.df <- pot.df %>% 
  bind_cols(as.data.frame(tmp)) %>% 
  mutate(
    Seven_or_More = seven_or_more,
    At_Least_One = 1 - Zero
  )

pot.df.long <- reshape2::melt(pot.df, id.vars = c("Pot", "Num_Tix", "year"))


ggplot(pot.df.long) +
  geom_line(aes(x = Num_Tix, y = value, color = variable)) +
  ggtitle("Probability of Number of Winners at a Given Number of Tickets Sold") +
  ylim(0,1) +
  xlab("Number of Tickets Sold") + ylab("Probability")


ggplot(pot.df.long) +
  scale_x_continuous(breaks = seq(0, 1500, by = 100)) +
  geom_line(aes(x = Pot, y = value, color = variable)) +
  ggtitle("Probability of Number of Winners at a Given Pot") +
  ylim(0,1) +
  xlab("Pot") + ylab("Probability")
```

Side note: because we're using the Poisson distribution, we know that the expected
number of winning tickets for a given pot is also given by
$\lambda_{Pot} = \frac{NumTix(Pot)}{292201338}$.

Let $P_{Pot}(i)$ denote the Poisson probability of $i$ **other** winners when the pot is of size $Pot$.
To calculate the expected value of the pot for **you**, then, we calculate:

$$E(Pot) = \frac{Pot}{1}*P_{Pot}(0) + \frac{Pot}{2}*P_{Pot}(1) + \frac{Pot}{3}*P_{Pot}(2) + ...$$

or

$$E(Pot) = Pot*\sum_{i=0}^{\infty} \frac{P_{Pot}(i)}{i+1}$$

or

$$E(Pot) = Pot*\sum_{i=0}^{\infty} \frac{\lambda_{Pot}^{i} e^{-\lambda_{Pot}}}{(i+1)i!} $$

or (changing indexes)

$$E(Pot) = \frac{Pot}{\lambda_{Pot}}*\sum_{i=1}^{\infty} \frac{\lambda_{Pot}^{i} e^{-\lambda_{Pot}}}{i!}$$

or

$$E(Pot) = \frac{Pot}{\lambda_{Pot}}(1-P_{Pot}(0))$$

or

$$E(Pot) = \frac{Pot*292201338}{NumTix(Pot)}(1 - e^{-NumTix(Pot)/292201338})$$

Note that this is
the expected value of the pot, not the pot's contribution to your expected
value from buying a ticket. Thus, we get something like this:

$$E_{Pot}(Ticket) = `r e` + \frac{E(Pot)}{292201338}$$

or

$$E_{Pot}(Ticket) = `r e` + \frac{Pot}{NumTix(Pot)}(1 - e^{-NumTix(Pot)/292201338})$$

```{r}
pot.df <- pot.df %>%
  mutate(
    Expected.Jackpot = map2_dbl(Pot, Num_Tix, ~ .x*1000000*(1 - exp(-.y/odds$one_over_p[1]))/.y),
    Expected = Expected.Jackpot + e,
    Naive.Expectation = e + Pot*1000000/odds$one_over_p[1]
  )

pot.df %>% 
  select(Pot, Expected.Jackpot, Expected, Naive.Expectation) %>% 
  reshape2::melt(id.vars = "Pot") %>% 
  ggplot() +
    scale_x_continuous(breaks = seq(0, 1500, by = 100)) +
    scale_y_continuous(breaks = 0.5*(0:10), minor_breaks = seq(0, 5, by = 0.1)) +
    geom_hline(yintercept = 2, linetype = "dashed", color = "black") +
    geom_line(aes(x = Pot, y = value, color = variable)) +
    ggtitle("Expectation at a Given Pot") +
    xlab("Pot") +
    ylab("Expectation") +
    theme(legend.position = 'top')
```



```{r eval=FALSE}
tix.lm <- lm(log(Num_Tix) ~ log(Pot) + I(log(Pot)^2) + jackpot_run + last_log_Pot + new_Pot,
             data = numbers)

par(mfrow = c(2,2))
plot(tix.lm)
summary(tix.lm)
par(mfrow = c(1,1))
shapiro.test(residuals(tix.lm))

numbers %>% 
  mutate(pred = exp(fitted(tix.lm))) %>% 
  ggplot(aes(x = Pot, y = Num_Tix, color = jackpot_run, group = jackpot_run)) +
  scale_y_log10() +
  scale_x_log10(breaks = c(100, 200, 300, 400, 500, 750, 1000, 1500), minor_breaks = NULL) +
  geom_point(shape = 1) +
  geom_line() +
  geom_line(aes(y = pred), linetype = "dotted") +
  ggtitle("Number of Tickets Sold by Jackpot Size") +
  xlab("Jackpot Size") +
  ylab("Number of Tickets Sold")
```
